---
title: "Homework 2 PSTAT 131"
author: "Idris Rasheed and Faraz Farooq"
date: "__Due on April 25th, 2017 at 11:59 pm__"
graphics: yes
geometry: margin=0.75in
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE, 
                      cache=TRUE, 
                      fig.width=5, 
                      fig.height=5,
                      fig.align='center')
indent1 = '    '      
indent2 = paste(rep(indent1, 2), collapse='')
solcode = TRUE
r = function(x, digits=2){ round(x, digits=digits) }
```

###1. 
     **Image compression with PCA**

    Bitmap image can be read in via the following command:

```{r readbitmap,indent=indent1}
#install.packages('bmp')
library(bmp)
library(dplyr)
img = read.bmp('image1.bmp')
img = t(img[ncol(img):1,])      # fix orientation
img = img - mean(img)           # subtract overall mean
```

    Plot the image in grayscale:

```{r plot-bitmap,indent=indent1}
gs = grey((0:255)/255)
image(img, asp=1, col=gs, yaxs='r', xaxt='n', yaxt='n')
```

###1a) 
```{r}
??bmp::xaxt
??bmp::yaxt
??bmp:image
```

###1b)
```{r}
pca_img <- prcomp(img)
str(pca_img)
names(pca_img)
```

###1c)
```{r}
X <- img
PHI <- pca_img$rotation
Z <- pca_img$x
(norm(Z-(X%*%PHI), type = "F"))^2
#This shoudld be zero but something is not working with our code
```

###1d)
```{r}
Q <- pca_img$rotation
(norm((t(Q)%*%Q)-diag(512), type = "F"))^2
```

###1e)
```{r}
new_img <- Z[,10:100]%*%t(PHI[,10:100])
image(new_img, asp=1, col=gs)
```

###1f)
```{r}
percent_var_expl <- pca_img$sdev^2 / sum(pca_img$sdev^2)
screeplot(pca_img, npcs = 10, type = "lines")
plot(percent_var_expl)
cumVar <- cumsum(pca_img$sdev^2 / sum(pca_img$sdev^2))
plot(cumVar)
summary(prcomp( .9*(pca_img$sdev^2 / sum(pca_img$sdev^2))))

```
From the plot's we can tell there are about 18 principal components needed to explain 90% of the variance

###3a)
```{r}
algae <- read.table('algaeBloom.txt',header=F,dec='.',
col.names=c('season','size','speed','mxPH','mnO2','Cl','NO3','NH4','oPO4',
'PO4','Chla','a1','a2','a3','a4','a5','a6','a7'),
na.strings=c('XXXXXXX'))
attach(algae)

sum(is.na(algae))

colSums(is.na(algae))

```
i)There are 33 total observations that contain missing values
ii)The only variables that contain missing values are mxPH (1), mn02(2), Cl(10), NO3(2), NH4(2),oPo4(2), Po4(2), Chla(12)

###3b)
```{r}
algae.del <- algae%>%filter(complete.cases(.))
algae.del
nrow(algae.del)
```
There are 184 observations in algae.del

###3c)
```{r}
algae.med <- algae %>%mutate_each(funs(ifelse(is.na(.), median(., na.rm=TRUE), .)))
nrow(algae.med)
algae.med[c(48,62,199),]
```
There are 200 observations

###3d)
```{r}
cor(algae.med, use= "complete.obs", method = "pearson")

prediction <- predict(lm(PO4~oPO4, data = algae.med))
prediction[28]

```
48.04407 is our value for the 28th observation


###4a)
```{r}
library(tree)
library(ISLR)
set.seed(1)
dim(algae.med)
train = sample(1:nrow(algae.med), 0.8*dim(algae.med)[1])
train
tree.algae.med = tree(lm(a1~ season + size + speed + mxPH +mnO2 + Cl + NO3 + NH4 +oPO4 + PO4 + Chla), data = algae.med, subset = train)
#Training error rate
algae.med.train= algae.med [train,]
a1.test = algae.med.train$a1
a1.pred = predict(tree.algae.med, algae.med.train)
error1 = table(a1.pred, a1.test)
error1
sum(diag(error1))/sum(error1)
1-sum(diag(error1))/sum(error1)

#Test error rate
algae.med.test = algae.med [-train,]
algae.med.test
a12.test = algae.med.test$a1
a12.pred = predict(tree.algae.med, algae.med.test)
error12 = table(a12.pred, a12.test)
error12
sum(diag(error12))/sum(error12)
1-sum(diag(error12))/sum(error12)


```

The training error rate is 90.625% and the test error rate is 82.5%

###4b)
```{r}
set.seed(2)
dim(algae.med)
train = sample(1:nrow(algae.med), 0.8*dim(algae.med)[1])
train
tree.algae.med = tree(lm(a1~ season + size + speed + mxPH +mnO2 + Cl + NO3 + NH4 +oPO4 + PO4 + Chla), data = algae.med, subset = train)
#Training error rate
algae.med.train= algae.med [train,]
a1.test = algae.med.train$a1
a1.pred = predict(tree.algae.med, algae.med.train)
a1.pred
error1 = table(a1.pred, a1.test)
error1
sum(diag(error1))/sum(error1)
1-sum(diag(error1))/sum(error1)

#Test error rate
algae.med.test = algae.med [-train,]
algae.med.test
a12.test = algae.med.test$a1
a12.pred = predict(tree.algae.med, algae.med.test)
a12.pred
error12 = table(a12.pred, a12.test)
error12

sum(diag(error12))/sum(error12)
1-sum(diag(error12))/sum(error12)

```
The training error rate went down to 86.25% and the test error rate went up to 90% as the seed increased our training error rate went down however the test error rate went up. It might be because the increase in the seed causes it to become more random. 

###5a)
```{r}
set.seed(1)
x<- cut((1:nrow(algae)), breaks=5, labels= FALSE)
sample(x)
```

###5b
```{r}
do.chunk <- function(chunkid, chunkdef, algae.med){ # function argument
train = (chunkdef != chunkid)
Xtr = algae.med[train,1:11] # get training set
Ytr = algae.med [train,12] # get true response values in trainig set
Xvl = algae.med [!train,1:11] # get validation set
Yvl = algae.med [!train,12] # get true response values in validation set
lm.a1 <- lm(a1~., data = algae.med [train,1:12])
predYtr = predict(lm.a1) # predict training values
predYvl = predict(lm.a1,Xvl) # predict validation values
data.frame(fold = chunkid,
train.error = mean((predYtr - Ytr)^2), # compute and store training error
val.error = mean((predYvl - Yvl)^2)) # compute and store test error
}
```

###5c)

```{r}

```




###6a)
```{r}

# Bootstrap 100 times
runs <- 1:100

# Lists to save the results

for (i in runs){
  print(i)
  sampleIndex <- sample(1:nrow(algae.med), size = nrow(algae.med), replace = TRUE)
  sampleData <- algae.med[sampleIndex, ]
  
}

mu = NULL #mean values of boostrap
for(draw in 1:100) mu = c(mu,mean(runs[sample(1:100,size=100,replace=T)]))


# Bind it into a nice data frame
algaedata <- cbind(runs, season, size, speed, mxPH, mnO2, Cl, NO3, NH4, oPO4, PO4, Chla, a1, a2, a3, a4, a5, a6, a7)
algdat <- as.data.frame(algaedata)

algaemodel <- lm(runs ~season+ size+ speed+ mxPH+ mnO2+ Cl+ NO3+ NH4+ oPO4+ PO4+ Chla+ a1+ a2+ a3+ a4+ a5+ a6+ a7, data=algdat)
summary(algaemodel)
```